---
title: "Class Size and Alumni Donation Analysis, part 2"
author: "Hart Williams"
date: "2026-09-08"
output:
  word_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Question 1.

# a. Continue with the same data from homework 1 and fit a simple linear regression on the data, where the alumni giving rate is the response variable \(Y\) of interest and the percentage of classes with fewer than 20 students is the predictor variable \(X\).
```{r}
alumni <- read.csv("alumni.csv")

model = lm(alumni_giving_rate ~ percent_of_classes_under_20, data = alumni)
model
summary(model)
```
# b. Repeat part a. But test whether the slope significantly differs from 1 at the \(\alpha = 0.1\) level.
```{r}
out <- summary(model)   # linear model
c0 <- 1                 # the hypothesized slope value (1)
b1 <- out$coefficients["percent_of_classes_under_20", "Estimate"]      # Estimated slope from regression
sb1 <- out$coefficients["percent_of_classes_under_20", "Std. Error"]   # Standard error of the slope
n <- nrow(alumni)   # number of observations

# The t-statistic measures how many standard errors b1 is away from c0
t_obs <- (b1 - c0) / sb1

# The p-value converts the t-statistic into a probability: the chance of seeing a t-statistic this extreme or more extreme (in either direction) if the null hypothesis (H0: b1 = c0) was actually true.
p_value <- 2 * pt(abs(t_obs), df = n - 2, lower.tail = FALSE)

round(c(t_obs = t_obs, p_value = p_value), 4)
```
- t_obs = -2.9836. This means the estimated slope is about 2.98 standard errors below 1.
- The p-value is 0.0045. This means that if the true slope was really 1, there would only be a 0.45% chance of observing a sample slope this far from 1 (in either direction) due to random sampling variability.
- Since the p-value (0.0045) is less than α = 0.10, we have sufficient evidence to conclude that the true slope differs from 1.

# c. What is the value of \(R^2\)? Please interpret. What is the correlation coefficient \(r\) between \(X\) and \(Y\), and what is the relationship between \(r\) and \(R^2\)?
```{r}
summary(model)$r.squared
r <- cor(alumni$percent_of_classes_under_20, alumni$alumni_giving_rate) 
r
r^2
```
- The R^2 value is ~41.69, meaning approximately 41.69% of the variation in alumni giving rate is explained by the percent of classes under 20 students. The remaining 58.31% is due to other factors not captured by the model.
- The correlation coefficient r = ~0.646, indicating a moderately strong, positive linear relationship between percent of classes under 20 and alumni giving rate; schools with a higher percentage of small classes tend to have higher alumni giving rates.
- r captures both the direction (positive/negative) and strength of the linear relationship, while R^2 squares that value to express how much of the variability in Y is explained by X.

# d. Using R, compute a 95% confidence interval for the mean giving rate of universities with \(X = 50\), and a 95% prediction interval for the giving rate of one university with \(X = 50\). Which interval is wider, and why are the two intervals answering different questions?
```{r}
new_data <- data.frame( 
   percent_of_classes_under_20 = 50 
 ) 

# Calculating confidence interval for new data 
predict(model, 
         newdata = new_data, 
         interval = "confidence", 
         level = 0.95) 

# Calculating prediction range for new data 
predict(model, 
         newdata = new_data, 
         interval = "prediction", 
         level = 0.95) 
```
- A 95% confidence interval for the mean giving rate at X=50 is (22.21%, 28.79%).
- A 95% prediction interval for one university at X=50 is (4.36%, 46,64%). 
- The prediction interval is wider, as while the confidence interval estimates where the average giving rate lies for all universities with X = 50, the prediction interval estimates where one specific university's giving rate might fall, which is less certain; individual universities have more variability than the estimated mean.

# e. Plot the training data (i.e., the data used to fit the model) with the fitted regression line and include a 95% (pointwise) confidence band for the mean responses. What do you observe about the confidence band at the point \(\left(\bar{X}, \bar{Y}\right)\)? Is it narrower or wider compared to the rest?
```{r}
# Create a grid of X values as a data.frame (needed for predict())
x_grid <- data.frame(
  percent_of_classes_under_20 = seq(min(alumni$percent_of_classes_under_20),
                                    max(alumni$percent_of_classes_under_20),
                                    length.out = 100)
)

# Get fitted values + confidence band
conf_band <- predict(model, newdata = x_grid, interval = "confidence", level = 0.95)

# Plot original data 
 plot(alumni$percent_of_classes_under_20, 
      alumni$alumni_giving_rate, 
      xlab = "Percent of Classes Under 20", 
      ylab = "Alumni Giving Rate", 
      main = "Regression with 95% Confidence Band") 
  
# Add fitted regression line 
 lines(x_grid$percent_of_classes_under_20, conf_band[, "fit"], lwd = 2) 
  
# Add confidence band 
lines(x_grid$percent_of_classes_under_20, conf_band[, "lwr"], lty = 2) 
lines(x_grid$percent_of_classes_under_20, conf_band[, "upr"], lty = 2) 

# (x_bar, y_bar)
x_bar = mean(alumni$percent_of_classes_under_20)
y_bar = mean(alumni$alumni_giving_rate)
points(x_bar, y_bar, col="red", pch=19)
```
The confidence band is narrowest at x_bar, and it gets wider as you move away from x_bar in either direction.

# f. Show the ANOVA table for this regression using anova(). From the R output only (no hand computation), identify the regression and error sums of squares and the F-statistic. What null hypothesis does this F-test address, and how does the result relate to your t-test in part a?
```{r}
anova(model)
```
- The regression sum of squares (SSR) is 3539.8, the error sum of squares (SSE) is 4951.7, and the F-statistic is 32.884.
- The F-test addresses H0: b1 = 0, the same null hypothesis as the t-test for the slope from summary(model). Their results agree exactly: F = t^2, and the two tests give identical p-values.

# g. A colleague asks you to predict the giving rate for a university with \(X = 95\). Using the range of \(X\) in the data, explain why this prediction needs a warning even though the software will still produce a number.
```{r}
# Check the range 
range(alumni$percent_of_classes_under_20) 
```
Since X = 95 is outside the observed range of the predictor, this prediction is an extrapolation. Even though R can produce a predicted value, it should be interpreted carefully because the linear relationship may not continue beyond the observed data.

# ------------------------------------------------

Question 2. 
# Assume mean function \(E\left(Y|X\right) = 10 + 5X\)

# a. Generate data with \(X \sim N\left(\mu = 2, \sigma = 0.1\right)\), sample size \(n = 100\), and error term \(\epsilon \sim N\left(\mu = 0, \sigma = 0.5\right)\).
```{r}
# same as HW1
set.seed(7052)

X <- rnorm(n = 100, mean = 2, sd = 0.1)
epsilon <- rnorm(n = 100, mean = 0, sd = 0.5)

Y <- 10 + 5*X + epsilon
```

# b. Fit a simple linear regression to the simulated data from part a. What is the estimated prediction equation? Report the estimated coefficients and their standard errors. Are they significant? Clearly write out the null and alternative hypotheses, observed t-statistic(s), p-value(s), and interpret the estimates and test results. What is the fitted model’s MSE?
```{r}
model = lm(Y ~ X)
summary(model)
mse = summary(model)$sigma^2
print(mse)
```
- The estimated prediction equation is Y_hat = 5.57X + 9.02.
- The estimated intercept is b0 = 9.02 with a standard error of SE(b0) = 0.83.
- The estimated slope is b1 = 5.57 with a standard error of SE(b1) = 0.42.

- For the slope, H0: b1=0 and H1: b1≠0.
- For the intercept, H0: b0=0 and H1: b0≠0.

- For the slope, t=13.39 on 98 df, with a p-value of ~0, or < 2*10^-16. This is below any conventional significance level, so we reject H0 and conclude there is a statistically significant linear relationship between X and Y.
- For every 1-unit increase in X, Y is predicted to increase by about 5.57 units on average.

- For the intercept, t=10.822 on 98 df, with a p-value of ~0, or < 2*10^-16. This  is below any conventional significance level, so we reject H0 and conclude that the intercept is statistically significant different from zero.
- The predicted value of Y when X=0 is about 9.02; however, this should be interpreted only as a component of the regression line, not as a meaningful prediction when Y=0.

- The fitted model's MSE is 0.20.

# c. Repeat part b, but re-simulate the data and change the error term to \(\epsilon \sim N\left(0, \sigma = 1\right)\).
```{r}
set.seed(7052)
X <- rnorm(n = 100, mean = 2, sd = 0.1)
epsilon <- rnorm(n = 100, mean = 0, sd = 1)
Y <- 10 + 5*X + epsilon

model = lm(Y ~ X)
summary(model)
mse = summary(model)$sigma^2
print(mse)
```
- The estimated prediction equation is Y_hat = 6.13X + 8.04.
- The estimated intercept is b0 = 8.04 with a standard error of SE(b0) = 1.67.
- The estimated slope is b1 = 6.13 with a standard error of SE(b1) = 0.83.

- For the slope, H0: b1=0 and H1: b1≠0.
- For the intercept, H0: b0=0 and H1: b0≠0.

- For the slope, t=7.38 on 98 df, with a p-value of ~0, or < 5.25*10^-11. This is below any conventional significance level, so we reject H0 and conclude there is a statistically significant linear relationship between X and Y.
- For every 1-unit increase in X, Y is predicted to increase by about 6.13 units on average.

- For the intercept, t=4.82 on 98 df, with a p-value of ~0, or < 5.16*10^-6. This  is below any conventional significance level, so we reject H0 and conclude that the intercept is statistically significant different from zero.
- The predicted value of Y when X=0 is about 8.04; however, this should be interpreted only as a component of the regression line, not as a meaningful prediction when Y=0.

- The fitted model's MSE is 0.81.

# d. Repeat parts a–c using \(n = 400\). What do you conclude? What is the effect on the model parameter estimates when the error variance gets smaller? What is the effect when the sample size gets bigger?
```{r}
set.seed(7052)
X <- rnorm(n = 400, mean = 2, sd = 0.1)
epsilon <- rnorm(n = 400, mean = 0, sd = 0.5)
Y <- 10 + 5*X + epsilon

print("Results for n=400, error sd=0.5")
model = lm(Y ~ X)
summary(model)
mse = summary(model)$sigma^2
print(mse)

print("Results for n=400, error sd=1")
epsilon <- rnorm(n = 400, mean = 0, sd = 1)
Y <- 10 + 5*X + epsilon
model = lm(Y ~ X)
summary(model)
mse = summary(model)$sigma^2
print(mse)
```
- In every scenario, the slope and intercept remain highly statistically significant, and the true relationship Y = 10 + 5X is strong enough to detect even with the added noise. However, the precision of estimates changes based on n and the error standard deviation.

- When the error standard deviation shrinks from 1 to 0.5, standard errors shrink roughly by half, and with less noise the point estimates land closer to the true values.

- When the sample size n goes from 100 to 400, standard errors shrink given more information, and likewise the point estimates get closer to the true values.

# e. What about the MSE from each model?
- When error standard deviation shrinks from 1 to 0.5, MSE is reduced to roughly one quarter of its original value. This is because MSE estimates error variance rather than error standard deviation; halving the error standard deviation should roughly quarter the MSE.

- Increasing the number of observations had little systematic effect on the MSE. This is expected given that MSE is an unbiased estimator of error variance, regardless of sample size.

# -----------------------------------------------
Question 3.

# a. For the first Question 2 setting, simulate 1,000 datasets; for each, fit the regression and save the estimated slope. Report the mean and the standard deviation of your 1,000 slopes. Is the mean close to the true \(\beta_1 = 5\)? How does the standard deviation compare with the Std. Error your single Question 2 fit reported — and what does that agreement tell you the standard error is?
```{r}
set.seed(7052)

n_sims <- 1000
n <- 100
slopes <- numeric(n_sims)

# Simulate 1000 datasets and save the estimated slope
for (i in 1:n_sims) {
  X <- rnorm(n, mean = 2, sd = 0.1)
  epsilon <- rnorm(n, mean = 0, sd = 0.5)
  Y <- 10 + 5 * X + epsilon
  fit <- lm(Y ~ X)
  slopes[i] <- coef(fit)["X"]
}

mean(slopes)   # average of the 1000 slope estimates
sd(slopes)     # spread of the 1000 slope estimates

# Compare to the single fit from the original Question 2 code
summary(model)$coefficients
```
- The mean slope is 5.03, which is essentially the true slope B1=5.
- The standard deviation of the slopes is ~0.51.
- The standard error for X in the single fitted model is ~0.42.

- The standard deviation of the 1,000 simulated slopes (0.51) is close to the standard error from the single regression fit (0.42), with both around 0.4-0.5. This general agreement supports the idea that the reported standard error is estimating the same thing as sd(slopes).

# b. Repeat part a with \(n = 400\) (same \(\sigma\)). Quadrupling the sample size multiplied the standard deviation of the slopes by about what factor?
```{r}
set.seed(7052)

n_sims <- 1000
n <- 400
slopes <- numeric(n_sims)

# Simulate 1000 datasets and save the estimated slope
for (i in 1:n_sims) {
  X <- rnorm(n, mean = 2, sd = 0.1)
  epsilon <- rnorm(n, mean = 0, sd = 0.5)
  Y <- 10 + 5 * X + epsilon
  fit <- lm(Y ~ X)
  slopes[i] <- coef(fit)["X"]
}

mean(slopes)   # average of the 1000 slope estimates
sd(slopes)     # spread of the 1000 slope estimates

# Compare to the single fit from the original Question 2 code
summary(model)$coefficients
```

Quadrupling the sample size multiplied the standard deviation of the slopes by a factor of about 0.5.

# c. In the part-a loop, also save both variance estimates from each fit: \(\mathrm{RSS}/(n-2)\) (the MSE) and \(\mathrm{RSS}/n\) (the maximum-likelihood estimate). Report the average of each across the 1,000 datasets and compare both to the true \(\sigma^2 = 0.25\). Which one is unbiased? For the other: is it too high or too low, by roughly what factor, and what do the “\(-2\)” degrees of freedom correspond to?
```{r}
set.seed(7052)

n_sims <- 1000
n <- 100
mse_var   <- numeric(n_sims)  # RSS/(n-2)
mle_var   <- numeric(n_sims)  # RSS/n

# Simulate 1000 datasets and save the MSE and MLE
for (i in 1:n_sims) {
  X <- rnorm(n, mean = 2, sd = 0.1)
  epsilon <- rnorm(n, mean = 0, sd = 0.5)
  Y <- 10 + 5 * X + epsilon
  fit <- lm(Y ~ X)
  
  RSS         <- sum(residuals(fit)^2)
  mse_var[i]  <- RSS / (n - 2)
  mle_var[i]  <- RSS / n
}

mean(mse_var)
mean(mle_var)
```
- The MSE is ~0.25 and the MLE is very slightly lower at ~0.245.
- The MSE is unbiased: its average across the 1,000 datasets should land right around 0.25, matching the true sigma^2 essentially exactly.
- The MLE is biased low; its average will be a bit below 0.25. For n=100, the expected shrinkage factor is (n-2)/2, which in this case is 98/100 = 0.98. In other words, the MLE is about 2% too low.
- The -2 degrees of freedom are "used up" by estimating the regression parameters (b0 and b1).




